\documentclass[conference]{IEEEtran}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{booktabs}
\usepackage{etoolbox}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta,fit,backgrounds}

% Horizontal justification for table notes
\patchcmd{\table}{\refstepcounter{table}}{\refstepcounter{table}}{}{}

\hyphenation{LEO-PGRL-LRFHSS}

\begin{document}

% ── Title ────────────────────────────────────────────────────────────────────
\title{PGRL-Assisted Uncertainty-Aware LR-FHSS Uplink Control for Direct-to-Satellite IoT}

\author{
  \IEEEauthorblockN{Z. D. Vanisherz}
  \IEEEauthorblockA{Independent Researcher\\
    Vanisherz@proton.me}
}

\maketitle

% ── Abstract ─────────────────────────────────────────────────────────────────
\begin{abstract}
We propose an uncertainty-aware Physics-Guided Residual Learning (PGRL) framework for Low-Earth-Orbit (LEO) IoT Long-Range Frequency-Hopping Spread Spectrum (LR-FHSS) control. A deterministic trajectory predictor is augmented with a calibrated log-variance head trained using Gaussian negative log-likelihood (NLL) while keeping the mean head frozen, so the point predictor is unchanged. The learned uncertainty is near-calibrated without post-hoc rescaling (best temperature $T=1.0$, Cov68/Cov95 $=0.713/0.947$). In a risk-aware guard-control ablation, calibrated uncertainty reduces an outage proxy from 5.0\% to 1.7\% with 13.8\% guard overhead, while deterministic position RMSE remains unchanged at 5.35~m. We additionally define a planned, not-yet-executed Semtech LR1121/USRP measurement interface. The contribution is thus calibrated uncertainty enabling risk-aware LR-FHSS/PGRL control, not improved deterministic trajectory accuracy.
\end{abstract}

\begin{IEEEkeywords}
Direct-to-Satellite IoT, LR-FHSS, Physics-Guided Residual Learning, calibrated uncertainty, risk-aware control, guard-band adaptation, LEO satellite
\end{IEEEkeywords}

% ── Introduction ─────────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:intro}

Direct-to-Satellite (D2S) IoT using Low-Earth Orbit (LEO) satellites enables connectivity for power-constrained end-devices in remote areas, but the high relative velocity between terminal and satellite---up to 8~km/s for LEO---induces Doppler shifts exceeding $\pm 30$~kHz and rapid Doppler-rate changes that stress the physical layer~\cite{hurn2023lrfhss}. Long-Range Frequency-Hopping Spread Spectrum (LR-FHSS), as standardized by Semtech in the AN1200.64 specification~\cite{semtech2023lrfhss}, achieves robust uplink by distributing transmit energy across a grid of narrow frequency bins, providing frequency diversity against Doppler uncertainty. However, LR-FHSS guard bands must accommodate worst-case timing and frequency uncertainty, imposing idle-time overhead that erodes the energy budget of battery-powered IoT nodes.

Existing approaches treat orbital prediction and physical-layer control separately: TLE-based propagators such as SGP4/SDP4~\cite{vallado2007fundamentals} provide deterministic orbital state but accumulate errors over prediction horizons beyond a few minutes, while LR-FHSS systems typically rely on wide guard bands or continuous ground-station feedback to track Doppler. We propose to close this loop with Physics-Guided Residual Learning (PGRL), which uses SGP4 as a physically-consistent anchor and augments a deterministic residual predictor with a \emph{calibrated} uncertainty head. Rather than improving the point prediction, the calibrated uncertainty lets the IoT terminal size its guard interval to the actual prediction risk: a risk-aware guard policy trades a small, bounded guard overhead for a substantial reduction in an outage proxy.

This paper's primary contributions are:
\begin{itemize}
  \item An uncertainty-aware PGRL residual learner with a calibrated log-variance output head, trained with Gaussian NLL while the deterministic mean head is frozen.
  \item A post-hoc temperature-calibration validation showing the learned uncertainty scale is already near-optimal ($T=1.0$, Cov68/Cov95 $=0.713/0.947$), i.e.\ no post-hoc rescaling is needed.
  \item A risk-aware LR-FHSS guard-adaptation ablation in which calibrated uncertainty reduces an outage proxy from 5.0\% to 1.7\% at 13.8\% guard overhead, with deterministic position RMSE unchanged at 5.35~m.
  \item A clearly scoped, planned (not-yet-executed) Semtech LR1121/USRP IQ measurement path.
\end{itemize}

\begin{figure}[t]
    \centering
    \resizebox{\linewidth}{!}{%
    \begin{tikzpicture}[
      font=\scriptsize, >=Latex, node distance=4mm and 5mm,
      box/.style={draw, rounded corners, align=center, minimum height=7mm, inner sep=2.5pt, fill=blue!4},
      ctrl/.style={draw, rounded corners, align=center, minimum height=7mm, inner sep=2.5pt, fill=orange!8},
      hw/.style={draw, dashed, rounded corners, align=center, minimum height=7mm, inner sep=2.5pt, fill=gray!8},
      ar/.style={->, thick}
    ]
    % Band 1: orbital prediction
    \node[box] (tle) {TLE source};
    \node[box, right=of tle] (sgp4) {SGP4/SDP4\\propagator};
    \node[box, right=of sgp4, fill=blue!10] (pgrl) {PGRL corrector\\frozen mean + log-var head};
    \node[box, right=of pgrl] (out) {\texttt{PGRLOutput}\\(mean, $\sigma_r$)};
    \draw[ar] (tle)--(sgp4);
    \draw[ar] (sgp4)--(pgrl);
    \draw[ar] (pgrl)--(out);
    % Band 2: risk-aware controller
    \node[ctrl, below=8mm of sgp4] (guard) {Risk-aware\\guard ($g{+}\alpha\sigma_r$)};
    \node[ctrl, right=of guard] (txt) {TX timing\\selection};
    \node[ctrl, right=of txt] (dop) {Doppler\\pre-comp};
    \draw[ar] (out.south) |- (dop.east);
    \draw[ar] (dop)--(txt);
    \draw[ar] (txt)--(guard);
    % Band 3: planned hardware
    \node[hw, below=8mm of guard] (lr) {Semtech LR1121 TX\\(SWDM001)};
    \node[hw, right=of lr] (usrp) {USRP B210\\IQ capture};
    \node[align=left, right=4mm of usrp, font=\scriptsize\itshape] (plan) {planned;\\not executed};
    \draw[ar] (guard.south) -- (lr.north);
    \draw[ar, dashed] (lr)--(usrp);
    \end{tikzpicture}}
    \caption{System architecture: TLE source $\rightarrow$ SGP4/SDP4 propagator $\rightarrow$ PGRL uncertainty-aware corrector (frozen mean head + calibrated log-variance head) $\rightarrow$ risk-aware uplink controller (guard-band, TX timing, Doppler pre-comp). The Semtech LR1121 / USRP B210 path is a planned, not-yet-executed measurement interface.}
    \label{fig:architecture}
\end{figure}

% ── Related Work ──────────────────────────────────────────────────────────────
\section{Related Work}
\label{sec:related}

\subsection{LR-FHSS for Direct-to-Satellite IoT}
LR-FHSS extends LoRaWAN for direct-to-satellite IoT by distributing transmit energy across a grid of narrow hopping bins, providing frequency diversity against the strong Doppler of LEO passes~\cite{semtech2023lrfhss,hurn2023lrfhss}. Prior LR-FHSS satellite-IoT studies focus largely on the \emph{receiver side}---synchronization, detection, packet delivery, and reliability recovery once a transmission is present. In contrast, this work targets \emph{terminal-side} control: how orbital-prediction uncertainty should shape uplink guard and timing decisions before transmission.

\subsection{ML for LEO Orbit Prediction}
Analytical TLE propagators such as SGP4/SDP4~\cite{hoots1980models,vallado2007fundamentals} are deterministic and lightweight but accumulate timing and Doppler error with prediction horizon and TLE age. Machine-learning orbit work increasingly treats such a propagator as a physics anchor and learns residual corrections, but the common objective is trajectory accuracy as an end in itself. This work instead uses SGP4-anchored residual learning to expose a \emph{calibrated uncertainty} that drives communication-side control, rather than to minimize point error.

\subsection{Gap and Positioning}
Table~\ref{tab:related} summarizes the positioning. Unlike receiver-side LR-FHSS studies, we do not target detection, packet delivery, or gateway scalability; unlike ML orbit-prediction work, we do not treat trajectory accuracy as the primary metric. Our contribution is the \emph{cross-layer connection}: converting SGP4-anchored, calibrated uncertainty into terminal-side risk-aware LR-FHSS guard control.

\begin{table}[htbp]
\small
\caption{Positioning Relative to Prior Art}
\label{tab:related}
\begin{tabular}{p{2.0cm}p{2.6cm}p{2.7cm}}
\toprule
Category & Typical focus & This work's difference \\
\midrule
LR-FHSS DtS analysis & Packet delivery / outage & No prediction-driven uplink control \\
LR-FHSS transceiver & Receiver sync / detection & Terminal-side pre-transmission control \\
ML orbit prediction & Trajectory accuracy & Calibrated uncertainty for control \\
\midrule
\textbf{This work} & Calibrated uncertainty $\rightarrow$ risk-aware guard & Cross-layer terminal design \\
\bottomrule
\end{tabular}
\end{table}

% ── Background and System Model ───────────────────────────────────────────────
\section{Background and System Model}
\label{sec:system}

\subsection{LEO D2S Channel Characteristics}
During a typical LEO satellite pass at 600~km altitude, a ground terminal experiences a maximum one-way Doppler of approximately 30~kHz at S-band (2~GHz) and a Doppler rate of $-50$ to $-150$~Hz/s near closest approach. The Doppler shift is modeled as
\begin{equation}
f_D(t) = -\frac{v_r(t)}{c}\,f_c
\label{eq:doppler}
\end{equation}
where $v_r(t)$ is the radial velocity (positive when satellite recedes) and $c$ is the speed of light. The residual CFO after pre-compensation is
\begin{equation}
f_{\mathrm{res}}(t) = f_D(t) - \hat{f}_D(t) + f_{\mathrm{osc}},
\label{eq:residual_cfo}
\end{equation}
where $\hat{f}_D$ is the predicted Doppler and $f_{\mathrm{osc}}$ is the terminal oscillator offset. In our LR-FHSS-inspired proxy, the available bandwidth is discretized into $N_f$ hopping bins. The bin width is treated as an evaluation parameter rather than a complete standard-compliant receiver model. Successful reception requires that $|f_{\mathrm{res}}|$ after pre-compensation be small relative to the bin width.

\subsection{SGP4/SDP4 Orbital Propagation}
SGP4/SDP4~\cite{hoots1980models} propagates Two-Line Element (TLE) sets to produce Cartesian position and velocity at any epoch. The algorithm is deterministic and computationally lightweight but accumulates RMS timing errors of 1--4~s and Doppler errors of $>5$~kHz over prediction horizons of 30--120~min, depending on TLE age. These errors are non-Gaussian and time-correlated, which limits the effectiveness of naive uncertainty estimation on top of SGP4 output.

\subsection{PGRL Architecture}
PGRL augments SGP4 with a SIREN (sinusoidal representation network) residual predictor operating on normalized orbital state from the SGP4 propagation. The network produces a $6$-dimensional deterministic mean output (position and velocity residuals); a parallel $6$-dimensional log-variance head $\log\sigma_i^2$ provides a per-axis uncertainty estimate. This is a heteroscedastic-uncertainty head, not a Bayesian posterior: there is no weight sampling, ensemble, or variational inference. The deterministic mean predictor is trained with a physics residual regularizer that encourages conservation of Keplerian energy $E=-\mu/(2a)$ and angular momentum $|\mathbf{h}|=\sqrt{\mu a(1-e^2)}$.

\emph{Uncertainty-head fine-tuning (Stage 3E).} Starting from the trained deterministic checkpoint, the mean head is \emph{frozen} and only the log-variance layer is trained with the Gaussian NLL objective
\begin{equation}
\mathcal{L}_{\text{NLL}} = \frac{1}{N}\sum_{n=1}^{N}\sum_{i=1}^{3}\left[
  \log\sigma_{n,i}^2 + \frac{(y_{n,i} - \mu_{n,i})^2}{\sigma_{n,i}^2}
\right],
\label{eq:nll}
\end{equation}
where $\mu_{n,i},\sigma_{n,i}$ are the (frozen) mean and learned per-axis standard deviation. The log-variance layer is initialized to $\log((10/\mathrm{DU})^2)$ (an initial per-axis $\sigma\approx 10$~m in reported metres, $\mathrm{DU}$ the distance unit) and converges to a mean axis $\sigma\approx 3.01$~m. Because the mean head is frozen, the point prediction---and therefore the deterministic accuracy---is unchanged by this stage. The learned uncertainty is thus a calibrated \emph{position-domain} ($\sigma$ in metres) estimate, not a directly trained timing (ms) or Doppler (Hz) uncertainty.

\emph{Coverage metric.} Calibration is assessed by Mahalanobis chi-square coverage: for each sample $z^2=\sum_i (e_i/\sigma_i)^2$ with per-axis error $e_i$. Under a calibrated $3$-DoF Gaussian, the 68\% and 95\% credible regions correspond to thresholds $z^2\le 3.506$ and $z^2\le 7.815$; the empirical coverage is the fraction of samples below each threshold (Sec.~\ref{sec:eval}).

% ── PGRL Orbital Predictor ────────────────────────────────────────────────────
\section{PGRL Orbital Predictor}
\label{sec:pgrl}

The PGRL predictor receives a TLE epoch and outputs a \texttt{PGRLOutput} dataclass containing:
\begin{itemize}
  \item \texttt{pass\_start\_error\_s}: predicted timing error relative to SGP4-estimated AOS;
  \item \texttt{doppler\_error\_Hz}: predicted residual Doppler after SGP4 compensation;
  \item \texttt{doppler\_rate\_error\_Hz\_per\_s}: predicted residual Doppler rate;
  \item \texttt{timing\_uncertainty\_sigma\_s}: one-sigma timing uncertainty;
  \item \texttt{doppler\_uncertainty\_sigma\_Hz}: one-sigma Doppler uncertainty;
  \item \texttt{link\_opportunity\_score}: composite score balancing Doppler, timing, and energy cost;
  \item \texttt{recommended\_guard\_time\_s}: minimum guard interval at $3.5\sigma$.
\end{itemize}

The trained uncertainty head emits a per-axis \emph{position-domain} standard deviation (in metres); the timing- and Doppler-domain $\sigma$ fields above are schema conveniences derived under the simulation mapping of Sec.~\ref{sec:uplink} and are not directly trained quantities. The uncertainty head maintains a near-calibrated scale without post-hoc rescaling (Sec.~\ref{sec:eval}); online/continual refinement is out of scope for this paper.

% ── Uplink Control ────────────────────────────────────────────────────────────
\section{LR-FHSS Uplink Control}
\label{sec:uplink}

\subsection{Risk-Aware Guard-Band Policy}
\label{sec:guard}
The calibrated position-domain uncertainty is used to size the guard margin to the actual prediction risk. Let $\sigma_r$ be the calibrated radial standard deviation (in metres) from the uncertainty head. The risk-aware guard is
\begin{equation}
g_{\text{uncertain}} = g_{\text{base}} + \alpha\,\sigma_r,
\label{eq:riskguard}
\end{equation}
where $\alpha\ge 0$ scales the uncertainty contribution. With $\alpha=0$ the policy reduces to the deterministic baseline (guard sized to the point prediction only); increasing $\alpha$ widens the guard in proportion to the predicted uncertainty. An \emph{outage proxy} counts the fraction of samples whose residual error exceeds the guard ($e_r>g_{\text{uncertain}}$); a \emph{success proxy} is its complement, and a risk-adjusted reward subtracts a small normalized guard-overhead penalty (Sec.~\ref{sec:eval}). We sweep $\alpha\in\{0.00,0.25,0.50,0.75,1.00\}$. All these quantities are control proxies, not measured packet-error rates.

\subsection{Doppler Pre-Compensation (Simulation/Proxy)}
As a separate control-framework component (a \emph{simulation/proxy assumption}, not part of the Stage~3E/3F/4 trained-model evidence), the \texttt{compensated\_tx\_frequency} function adjusts the nominal carrier $f_c$ using a predicted Doppler shift $\hat{f}_d$ and rate $\hat{f}_d'$:
\begin{equation}
f_{\text{TX}} = f_c - \hat{f}_d - \frac{\hat{f}_d'}{2}\,T_{\text{burst}},
\end{equation}
with $T_{\text{burst}}$ the burst duration. Under the simulation assumptions of Sec.~\ref{sec:eval}, tighter residual CFO keeps hop energy within an LR-FHSS bin; we report only proxy RF-quality indicators (EVM proxy, grid orthogonality) and do not measure PER.

\subsection{TX Timing Selection}
The \texttt{select\_tx\_time} function scores candidate transmission windows using a weighted utility:
\begin{equation}
U(t) = \alpha \cdot \text{link\_score}(t) - \beta \cdot |f_d'(t)| - \gamma \cdot E_{\text{idle}}(t)
\end{equation}
where $\alpha, \beta, \gamma$ are tunable weights. The optimizer selects the window with maximum $U(t)$ within a 6-hour planning horizon, balancing link quality, Doppler rate, and energy expenditure.

% ── Evaluation ────────────────────────────────────────────────────────────────
\section{Evaluation}
\label{sec:eval}

\subsection{Experimental Setup}
The uncertainty and control results (Stages 3E/3F/4) are produced by the released offline pipeline (\texttt{scripts/train\_uncertainty\_offline.py}, \texttt{calibrate\_uncertainty\_temperature.py}, \texttt{risk\_aware\_control\_ablation.py}) applied to the deterministic checkpoint, with seed~$=42$; trained checkpoints reside on the training machine and are not included in the repository. Reported metres use $e_m = e_{\text{norm}}\cdot\mathrm{DU}$. The risk-aware ablation (Eq.~\ref{eq:riskguard}) defines the outage proxy as the fraction of samples with $e_r>g_{\text{uncertain}}$, the success proxy as its complement, and the reward as $\text{success}-\lambda(\,\overline{g}-1)$ with energy weight $\lambda=0.1$ and $\overline{g}$ the guard normalized to the $\alpha{=}0$ baseline. As \emph{separate control-framework simulation proxies} (Monte Carlo, seed~42; not Stage~3E/3F/4 trained-model evidence) we additionally report an LR-FHSS 64-bin grid-orthogonality proxy and a QPSK EVM proxy at SNR$=40$~dB; these are RF-quality indicators only and are not LR-FHSS PER.

\subsection{Calibrated Uncertainty and Risk-Aware Control}
\input{tables/main_results}

\emph{Stage 3E (uncertainty-head fine-tuning).} With the mean head frozen, deterministic accuracy is unchanged: position RMSE remains 5.35~m and the fraction of predictions within 5~m remains 61.3\%. The log-variance head learns a metre-scale uncertainty, with the mean per-axis $\sigma$ moving from an initialized $\approx 10$~m to $\approx 3.01$~m, yielding non-trivial Mahalanobis coverage.

\emph{Stage 3F (temperature calibration).} Table~\ref{tab:tempsweep} sweeps a scalar temperature $T$ applied as $\log\sigma^2_T=\log\sigma^2+2\log T$. RMSE is invariant to $T$ (the mean is frozen). The best calibration error occurs at $T=1.0$ (Cov68/Cov95 $=0.713/0.947$, CalErr $=0.0331$), confirming that the learned uncertainty scale is already near-optimal \emph{without} post-hoc rescaling.

\emph{Stage 4 (risk-aware guard ablation).} Table~\ref{tab:risk} sweeps $\alpha$ in Eq.~\ref{eq:riskguard}. RMSE is fixed at 5.35~m across all $\alpha$, confirming the point predictor is untouched. Relative to the deterministic baseline ($\alpha{=}0$: outage 5.0\%, reward 0.9500), the uncertainty-aware policy at $\alpha{=}0.25$ reduces the outage proxy to 1.7\% (a $3\times$ reduction) and raises the success proxy to 0.9828 and the reward to 0.9690, at a guard overhead of $+13.8\%$. Larger $\alpha$ drives outage lower still but the risk-adjusted reward peaks at $\alpha{=}0.25$. The collision proxy is $0.000$ for all $\alpha$, so outage---not collision---is the binding robustness metric in this proxy setup; we therefore do not claim collision reduction. The separate simulation proxies give a QPSK EVM proxy of $\approx 67\%$ and a grid orthogonality of $0.706$ at 300~Hz residual CFO; these RF-quality proxies are not PER and are reported only to indicate that residual-CFO tracking would remain necessary for production operation.

\subsection{Validation Scope}
All reported results are simulation or proxy. Stages~3E/3F/4 are offline trained-model and control-proxy results; the grid/EVM rows are simulation/proxy assumptions, not trained-model evidence. This work does \emph{not} implement a standards-compliant LR-FHSS gateway and reports no LR-FHSS PER. The Semtech LR-FHSS TX and SDR IQ paths are defined but \emph{not executed}; no hardware packet delivery is claimed.

\subsection{Limitations}
This work does not claim improved deterministic trajectory RMSE from uncertainty fine-tuning; the mean predictor is frozen and remains at 5.35~m RMSE. The uncertainty head is a heteroscedastic log-variance output and is \emph{not} Bayesian in the posterior-sampling sense (no weight sampling, ensemble, or variational inference). The learned uncertainty is a position-domain ($\sigma$ in metres) quantity; the timing- and Doppler-domain figures are separate simulation/proxy assumptions and a modeling gap remains between position uncertainty and timing/Doppler control margins. The LR1121/USRP path is a planned measurement interface, not hardware packet-delivery validation. Outage, success, reward, and collision are control proxies rather than measured PER, and the collision proxy is $0.000$ for all $\alpha$, so no collision-reduction claim is made. There is no online or daily-TLE continual learning; the corrector is trained offline.

% ── Hardware Validation Plan ─────────────────────────────────────────────────
\section{Hardware Validation Plan}
\label{sec:hardware}

\subsection{Semtech LR1121 TX Path}
The Semtech LR1121/LR11xx dev-kit running SWDM001 firmware~\cite{swdm001} is the intended LR-FHSS transmitter for a future measurement stage. The \texttt{semtech\_validation/tx\_config\_from\_pgrl.py} script converts a \texttt{PGRLOutput} into a TX configuration JSON specifying center frequency, TX power, hopping-grid parameters, guard time, and Doppler pre-compensation flags. Only this config generation is exercised here (dry-run); no transmission has been captured or validated.

\subsection{SDR IQ Testbed}
A USRP B210 connected via coax (30~dB attenuator) to the Semtech TX forms the intended impairment testbed. The SDR is configured to capture IQ at 1~MHz bandwidth for a CFO estimator (phase-derivative method), an EVM proxy (QPSK constellation), and a waterfall visualizer. To date only the synthetic-IQ dry-run of these signal-processing blocks has been exercised; no live capture has been performed, and no hardware figure is reported.

% ── Conclusion ────────────────────────────────────────────────────────────────
\section{Conclusion}
\label{sec:conclusion}

This paper presented an uncertainty-aware PGRL framework for LR-FHSS LEO IoT control. Augmenting a frozen deterministic predictor with a calibrated log-variance head leaves the point prediction unchanged (position RMSE 5.35~m) while producing a near-calibrated uncertainty scale that needs no post-hoc rescaling ($T{=}1.0$, Cov68/Cov95 $=0.713/0.947$). In a risk-aware guard-control ablation, this calibrated uncertainty reduces an outage proxy from 5.0\% to 1.7\% at 13.8\% guard overhead. The contribution is therefore calibrated uncertainty enabling risk-aware LR-FHSS/PGRL control, not improved deterministic trajectory accuracy. A Semtech LR1121 + USRP B210 IQ measurement path is defined but not yet executed; hardware characterization is left to future work.

% ── References ───────────────────────────────────────────────────────────────
\bibliographystyle{IEEEtran}
\bibliography{refs}

\end{document}