\documentclass[conference]{IEEEtran}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{etoolbox}
\usepackage{enumitem}
\setlist[itemize]{leftmargin=1.2em,itemsep=1pt,topsep=2pt,parsep=0pt}

% Compact vertical spacing for a dense workshop paper (readability preserved)
\setlength{\textfloatsep}{6pt plus 2pt minus 2pt}
\setlength{\dbltextfloatsep}{7pt plus 2pt minus 2pt}
\setlength{\floatsep}{6pt plus 2pt minus 2pt}
\setlength{\intextsep}{6pt plus 2pt minus 2pt}
\setlength{\abovedisplayskip}{3pt plus 1pt minus 1pt}
\setlength{\belowdisplayskip}{3pt plus 1pt minus 1pt}
\setlength{\abovedisplayshortskip}{2pt}
\setlength{\belowdisplayshortskip}{2pt}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta,fit,backgrounds,calc}

% Horizontal justification for table notes
\patchcmd{\table}{\refstepcounter{table}}{\refstepcounter{table}}{}{}

\hyphenation{LEO-PGRL-LRFHSS}

\begin{document}

% ── Title ────────────────────────────────────────────────────────────────────
\title{Physics-First Evidence-Gated Uplink Control for LR-FHSS Direct-to-Satellite IoT}

\author{
  \IEEEauthorblockN{Author~1}
  \IEEEauthorblockA{Department of Example Engineering\\
    Example University\\
    City, Country\\
    author1@example.edu}
  \and
  \IEEEauthorblockN{Author~2}
  \IEEEauthorblockA{Department of Example Engineering\\
    Example University\\
    City, Country\\
    author2@example.edu}
  \and
  \IEEEauthorblockN{Author~3}
  \IEEEauthorblockA{Department of Example Engineering\\
    Example University\\
    City, Country\\
    author3@example.edu}
}

\maketitle

% ── Abstract ─────────────────────────────────────────────────────────────────
\begin{abstract}
Direct-to-Satellite (D2S) IoT terminals must pre-compensate large LEO Doppler
shifts before every LR-FHSS uplink from an on-board orbital prediction that is,
in practice, a \emph{stale} Two-Line Element (TLE) set propagated open-loop by
SGP4. A natural idea is to learn a data-driven residual correction on top of this
physics baseline. We test that idea on real Space-Track TLE history for two
BLACK~KITE LEO satellites (NORAD~66741, 68474) using strict chronological,
leakage-free splits, and find that a learned Doppler residual model does
\emph{not} improve over the open-loop SGP4 / stale-TLE baseline at any tested
staleness (8--168~h), either target-specific or cross-satellite: the inter-TLE
Doppler residual is essentially zero-mean and unpredictable. Rather than discard
learning, we make it conditional. We introduce an \emph{Evidence Gate} that
enables a learned corrector only when a chronological held-out validation window
shows it beats the physics baseline by a margin~$\gamma$, and otherwise falls
back to open-loop compensation. On the real BLACK~KITE data the gate correctly
\emph{closes} (learning disabled). In a controlled synthetic stress study we
characterise the regime in which the gate \emph{opens}: when a dominant
systematic drift exists, the gated corrector reduces simulated frequency error,
guard-band, and outage proxies, while a noise-dominated regime keeps the gate
shut. All ``reference'' Doppler is model-derived (SGP4 from real TLE), not
measured; no live-satellite, link-layer, or hardware-radiated result is claimed.
\end{abstract}

\begin{IEEEkeywords}
Direct-to-Satellite IoT, LR-FHSS, transmitter-side control, stale TLE, SGP4,
Doppler residual, evidence gate, safe-by-default learning, LEO satellite
\end{IEEEkeywords}

% ── Introduction ─────────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:intro}

Direct-to-Satellite (D2S) IoT using Low-Earth Orbit (LEO) satellites enables
connectivity for power-constrained end-devices in remote areas, but the high
relative velocity (several~km/s in LEO) induces Doppler shifts of tens of
kHz---order $20$~kHz at 868~MHz and roughly $50$~kHz at S-band, depending on
carrier frequency and pass geometry---together with rapid Doppler-rate changes
that stress the physical layer~\cite{hurn2023lrfhss}. Long-Range
Frequency-Hopping Spread Spectrum (LR-FHSS), characterized in Semtech's
AN1200.64 note~\cite{semtech2023lrfhss}, achieves robust uplink by distributing
transmit energy across a grid of narrow frequency bins, providing frequency
diversity against Doppler uncertainty. Because the sub-channels are narrow,
residual frequency error after pre-compensation is costly: it widens the guard
the terminal must reserve, increases the chance of missing a hop bin, and spends
airtime and energy that a battery-powered, unattended node cannot recover.

The terminal must size these margins \emph{before} transmitting, using only the
orbital information it holds on board---in practice a \emph{stale} TLE propagated
open-loop by SGP4/SDP4~\cite{vallado2007fundamentals}. This is the deployable
physics baseline. A tempting hypothesis is that a learned model can predict the
stale-TLE$\rightarrow$truth Doppler residual and shrink the error. The central
question of this paper is therefore not how to recover a waveform once it
exists, but a transmitter-side decision: \emph{when is it justified to learn a
correction on top of the physics prediction, and when is the safe physics
default the right choice?}

We expose a gap in the naive ``always learn'' answer. On real TLE history for two
BLACK~KITE satellites, a learned residual corrector does not beat the open-loop
SGP4 / stale-TLE baseline at any tested staleness; the residual between
successive TLE solutions is near zero-mean and unpredictable, so an always-on
learner would, if anything, \emph{increase} the frequency error it is meant to
reduce. Our stance is to treat learning as an option that must be \emph{earned}
on held-out evidence, with the physics model as the safe default. We formalise
this as an \emph{Evidence Gate}: a learned corrector is deployed only when a
chronological held-out validation window shows it beats the physics baseline by a
margin, and the controller otherwise uses open-loop compensation. On the real
data the gate closes; in a controlled synthetic stress study we show the gate
opens exactly when a dominant systematic drift makes the residual learnable, and
then reduces simulated frequency-error, guard, and outage proxies.

This paper's primary contributions are:
\begin{itemize}
  \item \emph{A physics-first, evidence-gated uplink controller.} A
  transmitter-side LR-FHSS control framework whose default is open-loop SGP4 /
  stale-TLE Doppler compensation and whose learned residual branch is enabled
  only by a held-out evidence test, integrated at the Doppler-precompensation
  point.
  \item \emph{A rigorous negative result on real data.} Using real Space-Track
  TLE history for two BLACK~KITE satellites and strict chronological splits, we
  show learned Doppler residual correction does not beat the open-loop baseline
  at 8--168~h staleness, both target-specific and cross-satellite.
  \item \emph{A controlled characterisation of when to learn.} A synthetic stress
  study spanning noise-dominated to systematic regimes, with $\gamma$ and
  validation-window sweeps, showing the gate closes on unlearnable residuals and
  opens on dominant systematic drift.
  \item \emph{An LR-FHSS guard / outage / energy proxy formulation.} A
  frequency-error-based set of control proxies (guard-band width, frequency-miss
  outage, and energy/overhead) used to evaluate the controller without claiming
  link-layer performance.
  \item \emph{Claim--evidence discipline and limitation-aware validation.} An
  explicit binding of every claim to a software-only artifact and to allowed
  wording, keeping model-derived results strictly separate from any measured or
  RF claim (of which there are none).
\end{itemize}

\begin{figure}[t]
    \centering
    \resizebox{0.90\columnwidth}{!}{%
    \begin{tikzpicture}[
        font=\footnotesize,
        >=Latex,
        block/.style={
            draw,
            rounded corners=3pt,
            align=center,
            minimum height=7.5mm,
            inner sep=3pt,
            fill=white
        },
        group/.style={
            draw,
            rounded corners=5pt,
            inner sep=4.5pt,
            fill=#1
        },
        arrow/.style={->, thick},
        keyarrow/.style={->, line width=2.2pt, green!45!black}
    ]

    % ===== Row 1: Prediction =====
    \node[block, fill=blue!10, minimum width=14mm] (tle)  at (0.4,0)     {Stale\\TLE};
    \node[block, minimum width=17mm]               (sgp4) at (2.9,0)     {SGP4\\open-loop};
    \node[block, fill=blue!8, minimum width=20mm]  (fphys)at (6.2,0.55)  {$f_{\mathrm{phys}}(t)$ baseline};
    \node[block, minimum width=20mm]               (fml)  at (6.2,-0.65) {$f_{\mathrm{ml}}=f_{\mathrm{phys}}+\hat r$};

    % ===== Row 2: Evidence Gate =====
    \node[block, fill=green!14, very thick, draw=green!45!black, minimum width=46mm]
        (gate) at (4.1,-2.5) {Evidence Gate\\$G=\mathbf{1}[\,\mathrm{MAE}_{\mathrm{ml}}(V)<\gamma\,\mathrm{MAE}_{\mathrm{phys}}(V)\,]$};

    % ===== Row 3: Terminal control =====
    \node[block, fill=orange!12, minimum width=22mm] (deploy) at (1.6,-4.5) {$\hat f=G f_{\mathrm{ml}}+(1{-}G)f_{\mathrm{phys}}$};
    \node[block, fill=orange!12, minimum width=22mm] (proxy)  at (6.6,-4.5) {guard / outage /\\energy proxy};

    % ===== Background group boxes =====
    \begin{scope}[on background layer]
        \node[group=blue!5,   fit=(tle)(sgp4)(fphys)(fml),
              label={[font=\footnotesize\bfseries, anchor=south west]north west:Prediction (model-derived)}] (G1) {};
        \node[group=green!7,  fit=(gate),
              label={[font=\footnotesize\bfseries, anchor=south west]north west:Evidence Gate (held-out validation)}] (G2) {};
        \node[group=orange!6, fit=(deploy)(proxy),
              label={[font=\footnotesize\bfseries, anchor=south west]north west:Terminal-side LR-FHSS Control}] (G3) {};
    \end{scope}

    % ===== arrows =====
    \draw[arrow] (tle) -- (sgp4);
    \draw[arrow] (sgp4.east) -- (fphys.west);
    \draw[arrow] (sgp4.east) -- (fml.west);
    \draw[arrow] (fphys.south) -- (gate.north);
    \draw[arrow] (fml.south)   -- (gate.north);
    \draw[keyarrow] (gate.south) -- node[right, font=\footnotesize\bfseries, green!45!black, xshift=1mm]
        {$G$} (deploy.north);
    \draw[arrow] (deploy.east) -- (proxy.west);

    % ===== Scope note =====
    \node[font=\scriptsize\itshape, text=gray!70]
        at ($(G3.south)+(0,-4mm)$)
        {Software-only / model-derived; no measured Doppler, no live-satellite, no RF claim.};

    \end{tikzpicture}}
    \caption{Physics-first evidence-gated uplink control. The open-loop SGP4 /
    stale-TLE Doppler $f_{\mathrm{phys}}$ is the default; a learned residual
    corrector $f_{\mathrm{ml}}=f_{\mathrm{phys}}+\hat r$ is deployed only when the
    Evidence Gate $G$ passes on a chronological held-out validation window. The
    deployed prediction feeds LR-FHSS guard / outage / energy control proxies.
    All Doppler is model-derived (SGP4 from real TLE);
    \texttt{reference\_is\_measured\_truth=false}.}
    \label{fig:architecture}
\end{figure}

% ── Related Work ──────────────────────────────────────────────────────────────
\section{Related Work}
\label{sec:related}

\subsection{LR-FHSS for Direct-to-Satellite IoT}
LR-FHSS extends LoRaWAN for direct-to-satellite IoT by distributing transmit
energy across a grid of narrow hopping bins, providing frequency diversity
against the strong Doppler of LEO passes~\cite{semtech2023lrfhss,boquet2021lrfhss}.
Link-level analyses quantify its behavior under LEO Doppler and large device
populations~\cite{ullah2022lrfhss,hurn2023lrfhss}, and energy studies characterize
the cost that headers, fragments, and idle margins impose on battery-powered
devices~\cite{sanchez2024energy}. A substantial body of work then improves what
happens \emph{after} a transmission exists: dedicated transceiver designs and
implementations~\cite{jung2023lrfhss,jung2025lrfhss}, header replication with
network coding~\cite{knop2024header}, asynchronous contention
resolution~\cite{santana2024acrda}, message replication for
reliability~\cite{rathi2024replication}, and probabilistic allocation of payload
code rate and header copies~\cite{farhat2025probabilistic}; real-world packet
traces further show that fragment recovery and demodulation are a dedicated
software-receiver problem~\cite{bukhari2023lrfhss}. These works study an existing
waveform's synchronization, detection, and delivery; the complementary
terminal-side question---how prediction quality should shape Doppler
pre-compensation \emph{before} transmission---is comparatively unexplored and is
our focus.

\subsection{ML for LEO Orbit Prediction}
Analytical TLE propagators such as
SGP4/SDP4~\cite{hoots1980models,vallado2007fundamentals,vallado2006revisiting}
are deterministic and lightweight but accumulate timing and Doppler error with
prediction horizon and TLE age. Machine-learning orbit work increasingly treats
such a propagator as a physics anchor and learns corrections on top of it:
Gaussian-process error correction~\cite{peng2019gp}, fusion of learned and
classical predictions~\cite{peng2021fusion}, differentiable SGP4
refinement~\cite{acciarini2025dsgp4}, physics-informed networks for satellite
state estimation~\cite{varey2024pinn}, and learned LEO orbit prediction with
exogenous variables~\cite{caldas2024leo}; see~\cite{caldas2024mlorbit} for a
survey. The dominant objective there is trajectory accuracy, with correctors
reported where they help. We instead ask the prior question for a terminal:
whether learning a Doppler residual over a stale-TLE baseline is justified by
held-out evidence \emph{at all}---and give a real-data case where it is not.

\subsection{LEO D2S Doppler and Uplink Control}
LEO non-terrestrial networking studies characterize the strong, time-varying
Doppler of LEO passes and motivate pre-compensation and timing control at the
terminal~\cite{3gpp2019tr38821,hurn2023lrfhss}, and uplink transmission-policy
work for LoRa-based D2S IoT optimizes when and how a terminal transmits using
pass geometry and link scores~\cite{alvarez2022uplink}. These policies typically
assume a deterministic prediction or a fixed conservative margin; the decision of
whether to trust a \emph{learned} correction is not itself treated as a gated,
evidence-driven choice. We make that choice explicit and safe-by-default.

\subsection{Gap and Positioning}
Receiver-side LR-FHSS work and terminal-side prediction-aware control are
complementary: the former recovers packets once transmitted, the latter decides
how transmissions are configured. Receiver/transceiver studies expose reliability
after transmission; D2S uplink policies expose a link score or pass window; and ML
orbit work exposes a trajectory estimate or residual corrector. The piece this
paper addresses is the transmitter-side decision under stale orbital
information---when to deploy a learned Doppler residual correction and when to keep
the physics default---bound to an explicit held-out evidence test.

% ── Background and System Model ───────────────────────────────────────────────
\section{Background and System Model}
\label{sec:system}

\subsection{LEO D2S Channel Characteristics}
In a LEO pass the one-way Doppler reaches tens of kHz (order $20$~kHz at 868~MHz,
$\sim$50~kHz at S-band), with the largest rate near closest approach and scaling
with carrier frequency and pass geometry. The Doppler shift is modeled as
\begin{equation}
f_D(t) = -\frac{v_r(t)}{c}\,f_c
\label{eq:doppler}
\end{equation}
where $v_r(t)$ is the radial velocity (positive when satellite recedes) and $c$
is the speed of light. The residual CFO after pre-compensation is
\begin{equation}
f_{\mathrm{res}}(t) = f_D(t) - \hat{f}_D(t) + f_{\mathrm{osc}},
\label{eq:residual_cfo}
\end{equation}
where $\hat{f}_D$ is the predicted Doppler and $f_{\mathrm{osc}}$ is the terminal
oscillator offset. We model reliable hop reception as $|f_{\mathrm{res}}|$ staying
small relative to a fixed frequency tolerance $F_{\mathrm{tol}}$; this is a control
proxy, not a standard-compliant receiver model.

\subsection{Stale-TLE Open-Loop Baseline and the Learned Residual}
\label{sec:baseline}
The terminal holds a TLE that is, in general, older than the present epoch.
Propagating this \emph{stale} TLE with SGP4~\cite{hoots1980models} gives the
deployable open-loop Doppler prediction $f_{\mathrm{phys}}(t)$. To study whether
this can be improved, we define a model-derived \emph{reference} Doppler
$D_{\mathrm{ref}}(t)$ as the SGP4 propagation of a \emph{later} TLE for the same
satellite at the same UTC. We emphasise that
\texttt{reference\_is\_measured\_truth = false}: $D_{\mathrm{ref}}$ is a newer
TLE solution, not an RF measurement of satellite Doppler. The quantity a learner
would have to predict is the inter-TLE residual
\begin{equation}
r(t) = D_{\mathrm{ref}}(t) - f_{\mathrm{phys}}(t),
\label{eq:residual}
\end{equation}
which encodes mean-element updates, drag-term changes, and numerical fit
differences between successive TLE solutions. A learned corrector forms
$f_{\mathrm{ml}}(t)=f_{\mathrm{phys}}(t)+\hat r(x_t)$ from a feature vector $x_t$
(TLE age, epoch gap, stale Doppler, orbital phase, geometry, and stale orbital
elements). All Doppler is computed at 868~MHz for a representative ground station
(24$^\circ$N, 121$^\circ$E).

% ── Method: Evidence Gate ─────────────────────────────────────────────────────
\section{Physics-First Evidence-Gated Control}
\label{sec:method}

\subsection{The Evidence Gate}
\label{sec:gate}
Let $V$ be a chronological \emph{validation} window, strictly later than the
training span and strictly earlier than the held-out test span, so that no future
TLE leaks into model selection. On $V$ we measure the mean absolute frequency
error of the physics baseline and of the learned corrector against the
model-derived reference,
\begin{align}
\mathrm{MAE}_{\mathrm{phys}}(V) &= \tfrac{1}{|V|}\textstyle\sum_{t\in V}\big|D_{\mathrm{ref}}(t)-f_{\mathrm{phys}}(t)\big|,\\
\mathrm{MAE}_{\mathrm{ml}}(V)   &= \tfrac{1}{|V|}\textstyle\sum_{t\in V}\big|D_{\mathrm{ref}}(t)-f_{\mathrm{ml}}(t)\big|.
\end{align}
The Evidence Gate enables learning only on a proven margin,
\begin{equation}
G = \mathbf{1}\!\left[\,\mathrm{MAE}_{\mathrm{ml}}(V) < \gamma\,\mathrm{MAE}_{\mathrm{phys}}(V)\,\right],
\qquad \gamma\in(0,1],
\label{eq:gate}
\end{equation}
and the deployed corrector defaults to physics otherwise,
\begin{equation}
\hat f(t) = G\,f_{\mathrm{ml}}(t) + (1-G)\,f_{\mathrm{phys}}(t).
\label{eq:deploy}
\end{equation}

\emph{Scope of the gate property (honest).} If $G=1$ then by construction
$\mathrm{MAE}_{\mathrm{ml}}(V) < \gamma\,\mathrm{MAE}_{\mathrm{phys}}(V) \le
\mathrm{MAE}_{\mathrm{phys}}(V)$, and if $G=0$ the deployed error equals the
baseline; hence $\mathrm{MAE}_{\hat f}(V)\le \mathrm{MAE}_{\mathrm{phys}}(V)$
\emph{on the validation window}. This is a validation-window property only, not a
held-out or distribution-shift bound: a gate-opened learner can still underperform
on unseen data under distribution shift. The threshold $\gamma$ trades safety
(small $\gamma$, rarely opens) against opportunism ($\gamma\!\to\!1$, opens on any
apparent gain, including noise); we therefore report the gate decision across
$\gamma$ and validation-window size and make no net-improvement claim.

\subsection{Statistical Role of the Gate}
\label{sec:stat}
The gate is a binary hypothesis test on the learned corrector, and it is useful
to name its two error modes. A \emph{false open} (Type-I) occurs when the residual
is noise-dominated but a favourable validation fluctuation drives
$\mathrm{MAE}_{\mathrm{ml}}(V)$ just below $\gamma\,\mathrm{MAE}_{\mathrm{phys}}(V)$,
so $G=1$ and the deployed pre-compensation inherits avoidable prediction noise
from a corrector that has nothing real to correct. A \emph{missed open} (Type-II)
occurs when a genuine systematic drift exists but the validation estimate does not
clear the margin, so $G=0$ and the terminal forgoes a chance to shrink its
frequency margin under the proxy of Sec.~\ref{sec:proxies}. The threshold
$\gamma$ is the conservative control knob between the two: a smaller $\gamma$
demands a wider proven margin, suppressing false opens at the cost of more missed
opens, while $\gamma\!\to\!1$ is opportunistic and will open on small, possibly
spurious validation gains. We use $\gamma=0.95$ as a conservative default, which
on the real BLACK~KITE data keeps the gate closed and on the synthetic regimes
opens only when the systematic margin is large (Sec.~\ref{sec:eval}). This framing
is descriptive of the decision rule's behaviour; it is not a performance promise
across deployment conditions.

\subsection{LR-FHSS Guard / Outage / Energy Proxies}
\label{sec:proxies}
Residual frequency error is the natural cross-layer quantity linking orbital
prediction to LR-FHSS terminal control: each hop occupies a narrow sub-channel,
so the deployed prediction error $e(t)=D_{\mathrm{ref}}(t)-\hat f(t)$ directly
sets how much frequency margin a transmission must reserve and how likely a hop is
to fall outside its intended bin. From $e(t)$ we derive three control proxies used
consistently across all experiments. The \emph{guard-band proxy}
$g = 2\,p_{99}(|e|)$ is a conservative frequency-margin proxy: twice the
99th-percentile absolute error covers the bulk of the error distribution while
remaining a percentile statistic rather than an absolute extremum. The
\emph{outage proxy} $\rho = \Pr[\,|e| > F_{\mathrm{tol}}\,]$, with a fixed
bin-tolerance $F_{\mathrm{tol}}=500$~Hz, represents the frequency-miss probability
under a simplified bin-tolerance model. The \emph{energy/overhead proxy} is
monotone in the reserved margin and in the repeated-transmission pressure that
missed hops induce; a convenient dimensionless form is
\begin{equation}
E_{\mathrm{proxy}} \;\propto\; \Big(1 + \alpha_g\,\tfrac{g}{B}\Big)\,(1 + \rho),
\label{eq:eproxy}
\end{equation}
where $B$ is the hopping bandwidth and $\alpha_g\!\ge\!0$ weights reserved guard
relative to retransmission pressure: the first factor charges for idle guard, the
second for the airtime that frequency misses tend to provoke. Equation~\eqref{eq:eproxy}
is an illustrative overhead \emph{proxy}, not a measured energy model and not a
packet-delivery model. All three are control-layer proxies for terminal margin
sizing; they are explicitly \emph{not} packet-error, bit-error, or link-budget
measurements.

\subsection{Datasets and Splits}
\label{sec:data}
Real evaluation uses Space-Track TLE history for two BLACK~KITE LEO satellites:
BLACK~KITE-1 (NORAD~66741, 415 records, 2025-12-18 to 2026-06-12, median
inter-TLE gap $6.3$~h) and BLACK~KITE-2 (NORAD~68474, 184 records, 2026-04-13 to
2026-06-13, median gap $6.4$~h). For each reference (newer) TLE, an older TLE
inside a target staleness band is paired with it, propagated open-loop, and
compared to the reference at 24 UTC samples over one orbital period; pairs with a
manoeuvre/bad-fit outlier are rejected. Splits are chronological by reference
epoch (60/20/20 train/validation/test), so the stale TLE in any test pair is
always available before the test reference epoch. The synthetic stress study
(Sec.~\ref{sec:eval-stress}) is a controlled simulation and uses no Space-Track
data.

\emph{Deployment recipe.} At control time the terminal propagates the stale TLE to
$f_{\mathrm{phys}}$, forms $f_{\mathrm{ml}}=f_{\mathrm{phys}}+\hat r(x_t)$,
evaluates $\mathrm{MAE}_{\mathrm{phys}}(V),\mathrm{MAE}_{\mathrm{ml}}(V)$ on $V$,
sets $G$ (Eq.~\eqref{eq:gate}) and deploys $\hat f$ (Eq.~\eqref{eq:deploy}), then
sizes the guard from $g=2\,p_{99}(|e|)$. It needs no receiver feedback and is
recomputed when a fresh TLE arrives.

% ── Evaluation ────────────────────────────────────────────────────────────────
\section{Evaluation}
\label{sec:eval}

\subsection{Experimental Setup}
Doppler is computed with python-SGP4 at 868~MHz for the representative ground
station of Sec.~\ref{sec:baseline}. The 868~MHz carrier is used as a
representative value for Doppler scaling and reproducibility; the controller is
frequency-parametric (Eq.~\eqref{eq:doppler}), so the same formulation applies at
other carriers by rescaling $f_c$, and no deployment-specific regulatory
assumption is made here. Learned candidates are ridge, random forest,
gradient boosting, and a small MLP; the learned model is selected on the
chronological validation segment and the gate is evaluated once on the held-out
test segment. A window is counted as supported only if the validation-selected
learned model beats the zero-residual stale-TLE baseline on held-out test MAE by
a clear margin and also improves RMSE. All results in this section are
software-only and model-derived; no hardware, RF, UART, replay, or measured
Doppler is involved.

\subsection{Real BLACK~KITE Negative Result}
\label{sec:eval-real}
Table~\ref{tab:bk} reports the held-out test MAE of the validation-selected
learned model against the zero-residual stale-TLE baseline. The learned residual
is worse than the baseline at \emph{every} tested staleness, both target-specific
(BK1, 8--168~h) and cross-satellite (BK1$\rightarrow$BK2), so the gate of
Eq.~\eqref{eq:gate} \emph{closes} in all cases. The held-out residual is
near zero-mean (e.g.\ BK1 test residual mean $|{\cdot}|\!\lesssim\!0.1$~Hz across
windows) with a spread that grows with staleness but no learnable structure: the
inter-TLE refresh ($\sim$6~h median) leaves little to correct at short staleness,
and BK1 manoeuvre/orbit-determination outliers make cross-satellite transfer
actively harmful. This is the paper's central real-data finding: on these
satellites, learning does not help, and the safe physics default is correct.

The near-zero-mean, low-structure residual is \emph{consistent with} the way
successive TLE solutions are produced. The inter-TLE Doppler residual may arise
from orbit-determination updates between fits, drag-model mismatch over the
propagation gap, tracking/fit noise, and occasional maneuver or bad-fit events;
these effects are largely idiosyncratic to each fit and are not exposed as
predictable terminal-side features in our input set (TLE age, epoch gap, stale
Doppler, orbital phase, and stale elements). We therefore read the negative
result as evidence that, for these satellites and features, the residual carries
little learnable signal---not as a claim about the optimality of any propagator.
Read constructively, this is a rigorous real-data case showing why always-on
residual learning is unsafe for a terminal that cannot verify the correction
before transmitting, and therefore why an evidence gate---rather than an
unconditional learned corrector---is the appropriate design. We call this a
\emph{weakly structured inter-TLE residual} for these satellites and features; we
do not assume it is white for all satellites, since other objects, longer
staleness, or richer features could expose learnable structure---precisely the
case the gate is built to admit when evidence supports it.

\begin{table}[t]
\centering
\small
\caption{Real BLACK~KITE held-out test MAE [Hz @ 868~MHz]: validation-selected
learned model vs.\ zero-residual stale-TLE baseline. $\Delta\%$ is the learned
model's change relative to the baseline (negative = worse). Gate closes in every
row. \texttt{reference\_is\_measured\_truth=false}.}
\label{tab:bk}
\begin{tabular}{llccr c}
\toprule
Setting & Stale & Baseline & Learned & $\Delta\%$ & Gate \\
\midrule
BK1 target & 8\,h   & 0.2430 & 0.3501 & $-44.1$ & closed \\
BK1 target & 24\,h  & 0.8161 & 0.9109 & $-11.6$ & closed \\
BK1 target & 48\,h  & 1.9433 & 2.8608 & $-47.2$ & closed \\
BK1 target & 72\,h  & 4.8947 & 5.9092 & $-20.7$ & closed \\
BK1 target & 96\,h  & 10.1153& 11.7663& $-16.3$ & closed \\
BK1 target & 168\,h & 26.9243& 45.2629& $-68.1$ & closed \\
\midrule
BK1$\to$BK2 & 8\,h  & 0.1877 & 0.3261 & $-73.7$ & closed \\
BK1$\to$BK2 & 24\,h & 0.4969 & 1.8639 & $-275.1$& closed \\
BK1$\to$BK2 & 48\,h & 2.4092 & 2.8458 & $-18.1$ & closed \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Controlled Synthetic Stress: When the Gate Opens}
\label{sec:eval-stress}
To characterise the gate mechanism we run a controlled synthetic simulation
(no Space-Track data, no measurement): the residual is generated as
$r = A\,s(x) + \mathcal{N}(0,\sigma^2)$ with a fixed nonlinear systematic
function $s(\cdot)$, and $(A,\sigma)$ chosen per regime to bracket gate
behaviour. Table~\ref{tab:stress} summarises three regimes. In the
noise-dominated \texttt{fresh\_low\_residual} regime---which matches the real
BLACK~KITE residual structure---the learned model cannot beat the baseline and
the gate closes. In the \texttt{extreme\_systematic} regime a dominant learnable
drift exists; the learned model cuts test MAE from $1859$ to $100$~Hz, the
guard-band proxy from $14{,}120$ to $950$~Hz, and the outage proxy from $0.814$
to $0.008$, and the gate opens. The \texttt{moderate\_staleness} regime is
marginal: the learned model is only $\sim$5\% better on test, so at the
recommended $\gamma=0.95$ the gate stays closed, trading a small missed
opportunity for not enabling a learner that has not decisively proven itself.

Interpreted through the proxy model, reducing the guard-band proxy from
$\approx 14.1$~kHz to $\approx 950$~Hz in the systematic regime corresponds to a
much smaller frequency margin the terminal would have to reserve, which under the
proxy can translate into lower conservative margin and overhead in terminal-side
control. We stress that this is an energy/overhead \emph{proxy} derived from the
deployed frequency error, not a measured power saving, a gateway receiver
search-window measurement, or a link-layer result; and that this benefit appears
only in the controlled synthetic regime, which the real BLACK~KITE data does not
exhibit.

\begin{table}[t]
\centering
\footnotesize
\setlength{\tabcolsep}{3.5pt}
\caption{Controlled synthetic stress (NOT real BLACK~KITE evidence). Held-out
test MAE [Hz], guard-band proxy $g=2\,p_{99}|e|$ [Hz], and outage proxy
$\Pr[|e|>500\,\mathrm{Hz}]$; gate at $\gamma=0.95$. Regimes: fresh
(noise-dominated), moderate (marginal), extreme (systematic).}
\label{tab:stress}
\begin{tabular}{lccccc}
\toprule
Regime & Base & ML & Gate & guard $b{\to}$ML & outage $b{\to}$ML \\
\midrule
fresh    & 9.56 & 9.61 & closed & 62$\to$61 & 0.000$\to$0.000 \\
moderate & 63.17 & 60.18 & closed & 412$\to$387 & 0.000$\to$0.000 \\
extreme  & 1859.30 & 99.75 & open & 14120$\to$950 & 0.814$\to$0.008 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Threshold and Validation-Window Sensitivity}
\label{sec:eval-sens}
Table~\ref{tab:gamma} sweeps the gate decision over $\gamma$ and over the
validation-window size. The two decisive regimes are insensitive to both: the
noise-dominated regime stays closed for every $\gamma\in[0.90,1.00]$ and every
validation size from $100$ to $3000$, and the strongly-systematic regime stays
open throughout. The threshold only matters at the boundary: the marginal regime
flips closed$\to$open between $\gamma=0.95$ and $\gamma=0.99$. Very small
validation windows ($n\!\approx\!100$) can mis-decide \emph{near} the threshold,
which is a property of any held-out estimate, not of the gate rule; the decision
is stable once the window is not tiny. The real BK1/BK2 validation segments are
well above this danger zone and sit deep in the noise-dominated regime, so the
real-data gate decision (closed) is unambiguous. We adopt $\gamma=0.95$ as a
conservative default.

\begin{table}[t]
\centering
\footnotesize
\setlength{\tabcolsep}{4pt}
\caption{Synthetic gate decision vs.\ threshold $\gamma$ (validation $n=3000$) and
vs.\ validation size (at the regime extremes). ``OPEN'' enables the learned
corrector.}
\label{tab:gamma}
\begin{tabular}{lcccc}
\toprule
Regime / window & $\gamma{=}.90$ & $\gamma{=}.95$ & $\gamma{=}.99$ & $\gamma{=}1.0$ \\
\midrule
fresh, $n{=}3000$    & closed & closed & closed & closed \\
moderate, $n{=}3000$ & closed & closed & OPEN & OPEN \\
extreme, $n{=}3000$  & OPEN & OPEN & OPEN & OPEN \\
\midrule
fresh, any $n$   & closed & closed & closed & closed \\
extreme, any $n$ & OPEN & OPEN & OPEN & OPEN \\
\bottomrule
\end{tabular}
\par\vspace{2pt}\raggedright\footnotesize\textit{``any $n$'': stable across
validation size $n\in\{100,300,1000,3000\}$.}
\end{table}

\subsection{Implications for Terminal Control}
\label{sec:implications}
Reading Tables~\ref{tab:bk}--\ref{tab:gamma} together yields a consistent control
rule. In noise-dominated regimes---which the real BLACK~KITE data exhibits---the
gate closes and the correct terminal action is to keep the open-loop SGP4 /
stale-TLE baseline; an always-on learner would add noise to the
pre-compensation (the false-open mode of Sec.~\ref{sec:stat}). In the systematic
synthetic regime the gate opens and the learned branch shrinks the guard-band
proxy from $\approx 14.1$~kHz to $\approx 950$~Hz and the outage proxy from
$0.814$ to $0.008$, so a terminal that has logged such evidence can reserve a
much smaller frequency margin. The moderate regime is exactly where the threshold
matters: a $\sim$5\% validation gain is not enough to open at $\gamma=0.95$,
deferring to physics rather than acting on a marginal signal. The
validation-window sweep shows the decision is statistically reliable once the
window is not tiny, so the gate's reliability is itself a function of how much
chronological evidence the terminal has accumulated. The control behaviour is
therefore data-dependent by construction: the same controller keeps physics where
learning is unjustified and admits learning only where the evidence is strong.

\subsection{Design Implications}
\label{sec:design}
Three practical rules follow. First, for terminals with fresh or frequently
refreshed TLEs---as in both BLACK~KITE feeds ($\sim$6~h refresh)---the open-loop
SGP4 prediction is already hard to beat, and a learned residual should not be
deployed unconditionally. Second, for stale or systematically biased deployments
the residual \emph{may} carry learnable structure, but it should be enabled only
after chronological validation demonstrates that structure, never by default.
Third, the controller should be deployed with evidence logging and explicit gate
decisions, so that the learned branch is auditable and reversible rather than
always-on. These are practical design rules grounded in the present software-only
study, not a universal performance claim.

% ── Limitations and Future Work ──────────────────────────────────────────────
\section{Limitations and Future Work}
\label{sec:limits}

This study is software-only and model-derived; we state its scope explicitly.
\begin{itemize}
\item \emph{Model-derived reference only.} All ``reference'' Doppler is SGP4 from
  real TLE (\texttt{reference\_is\_measured\_truth=false}); there is no measured
  Doppler truth.
\item \emph{No live or link-layer results.} No live-satellite contact or tracking,
  no standards-compliant LR-FHSS decoding, and no packet-error rate (PER),
  bit-error rate (BER), cyclic redundancy check (CRC), packet-delivery ratio
  (PDR), or gateway acknowledgement.
\item \emph{Negative result is not universal.} The real-data finding holds for two
  BLACK~KITE satellites and our feature set; it is not a claim for all satellites
  or all staleness/feature regimes.
\item \emph{Synthetic benefit is not real evidence.} The positive learning result
  appears only in a controlled simulation (Sec.~\ref{sec:eval-stress}) and must
  not be read as real-satellite evidence.
\item \emph{Gate scope.} The Evidence Gate provides a validation-window property
  (Sec.~\ref{sec:gate}), not a held-out or distribution-shift safety property.
\item \emph{Proxies only.} Guard, outage, and energy are control proxies, not
  link-budget or packet-delivery measurements.
\end{itemize}

\emph{Hardware-in-the-loop attempt.} A supervised Mac conducted hardware-in-the-loop
attempt was carried out and \emph{halted as inconclusive}: the physical bench
lacked an SMA/u.FL conducted RF path and a fixed attenuator between the LR1121
transmit output and the USRP~B210 receive input (the boards were only co-located,
with a LoRa antenna unsuitable as conducted evidence), so a TX-ON capture was not
distinguishable from the TX-OFF baseline. We therefore claim \emph{no} conducted
RF signal presence and no hardware validation from this attempt. A valid future
run requires a proper cabled path---LR1121 RF output $\rightarrow$ coax
$\rightarrow$ fixed attenuator $\rightarrow$ B210 RX---after which an IQ-level
TX-ON/OFF signal-presence diagnostic (and, separately, a gateway-class decoded
receiver for any link-layer metric) becomes possible. Closing this conducted-IQ
and receiver-in-the-loop gap is the primary future work.

% ── Conclusion ────────────────────────────────────────────────────────────────
\section{Conclusion}
\label{sec:conclusion}

We presented a physics-first, evidence-gated controller for transmitter-side
LR-FHSS Doppler pre-compensation under stale orbital information: open-loop SGP4 /
stale-TLE prediction is the safe default, and a learned residual is deployed only
when a chronological held-out window shows it beats the baseline by a margin. On
real BLACK~KITE TLE history the learned correction never beats the baseline, so
the gate closes; in a controlled synthetic study the gate opens and shrinks
simulated frequency-error, guard, and outage proxies only when a dominant
systematic drift exists. The outcome is a safe-by-default controller whose learned
component is enabled only on validation evidence. All results are software-only and
model-derived; closing the conducted-IQ and receiver-in-the-loop gap with a
properly cabled, attenuated bench is future work.

% ── References ───────────────────────────────────────────────────────────────
\bibliographystyle{IEEEtran}
\bibliography{refs}

\end{document}
